Nathan Greenslit Data Analysis Project

### 2023-04-20 code/analysis_code/stats/3_correlation.qmd
edits

```{r}
library(tidyverse)
library(here)
library(Hmisc)#Cor Tables and Lag
library(corrplot) #Cor Plots
library(lubridate) #For dates
```

@@

### Log10 Transform non-normally distributed data and Remove Non-Parametric Variables

<!--# If we are using a nonparametric correlation (i.e. Spearman) we do not log-transform data -->

Distributions can be found in `1_distribution.qmd`

```{r}
#Blind Oso
#bo<-
  #bo %>%
 # mutate(amm = log10(amm), #Log-transforms data
        # din_dip = log10(din_dip),
        # doc = log10(doc),
        # dust = log10(dust),
         #chl = log10(chl)) %>%
 # select(!c(secchi, tn, tdn, don)) #Removes non-parametric data

#Canals
#c2<- 
  #c2 %>%
  #mutate(orthop = log10(orthop),
         #dust = log10(dust)) %>%
  #select(!c(secchi, nn, sil, doc,toc, tn))

#Gulf
#rd<- 
 # rd %>%
 # mutate(doc = log10(doc),
      #   toc = log10(toc),
        # dust = log10(dust)) %>%
  #select(!c(sal, do_mgl, secchi, nn, tn))
```

@@

```{r, include = FALSE}
b<- cor(bo, method = "spearman") #Default method = pearson
round(b,2) #Round correlation coeff to "x" digits

c<- cor(c2, method = "spearman")
round(c,2)

r<- cor(rd, method = "spearman")
round(r,2)
```

#### Make Function to Format

Simple function for formatting a **correlation matrix** into a table with 4 columns containing : Column 1 : row names (variable 1 for the correlation test), Column 2 : column names (variable 2 for the correlation test), Column 3 : the **correlation coefficients**, Column 4 : the **p-values** of the correlations

```{r}
flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut])}
```

#### Make Matrix

The function **rcorr()** \[in **Hmisc** package\] can be used to compute the **significance levels** for **pearson** and **spearman correlations**. It returns both the correlation coefficients and the p-value of the correlation for all possible pairs of columns in the data table.

```{r}
bclean<- rcorr(as.matrix(bo))
flattenCorrMatrix(bclean$r, bclean$P)

cclean<- rcorr(as.matrix(c2))
flattenCorrMatrix(cclean$r, cclean$P)

rclean<- rcorr(as.matrix(rd))
flattenCorrMatrix(rclean$r, rclean$P)
```

@@

<!--# This is using Spearman's with raw data -->

```{r}
corrplot(b, type = "upper", 
         tl.col = "black", tl.srt = 45,
         title = "Blind Oso",  mar=c(0,0,1,0)) #Adds and lowers title

corrplot(c, type = "upper", 
         tl.col = "black", tl.srt = 45,
         title = "Canals", mar=c(0,0,1,0))

corrplot(r, type = "upper", 
         tl.col = "black", tl.srt = 45, mar=c(0,0,1,0))
         title(main = "Gulf" )
```

>>>>

```{r}
library(tidyverse)
library(here)
library(Hmisc)#Cor Tables and Lag
library(corrplot) #Cor Plots
library(lubridate) #For dates
library(ggpubr) #Plot stacking
```

@@

#### Make  datasets for MADA analysis

```{r}
b1<- 
  bo %>%
  select(temp, sal, nn, orthop, don, doc, tdn, chl, copies_mL, dust)

c1<- 
  c2 %>%
  select(temp, sal, nn, orthop, don, doc, tdn, chl, copies_mL, dust)

r1<- 
  rd %>%
  select(temp, sal, nn, orthop, don, doc, tdn, chl, copies_mL, dust)
```

@@

```{r, include = FALSE}
b2<- cor(b1, method = "spearman") 
round(b1,2) #Round correlation coeff to "x" digits

c2<- cor(c1, method = "spearman")
round(c1,2)

r2<- cor(r1, method = "spearman")
round(r1,2)
```

@@

```{r}
par(mfrow=c(1,3))

#Non-Parametric Data
 corrplot(b2, type = "upper", 
         tl.col = "black", tl.srt = 90,
         title = "Blind Oso",  mar=c(0,0,1,0)) #Adds and lowers title

corrplot(c2, type = "upper", 
         tl.col = "black", tl.srt = 90,
         title = "Canals", mar=c(0,0,1,0))

corrplot(r2, type = "upper", 
         tl.col = "black", tl.srt = 90, mar=c(0,0,1,0))
         title(main = "Gulf" )
```

#####

### 2023-04-20 products/manuscript/Manuscript.qmd
edits

The Sahara Desert is a significant source of atmospheric dust, eliciting an estimated one billion metric tons each year \[@dalmeidaModelSaharanDust1986\]. Annually, plumes of this dust travel across the Atlantic via easterly trade winds to be deposited in the surface waters of the Atlantic, Caribbean, and Gulf of Mexico. These dust events are highly episodic, occurring 3- 4 times a year typically in the summer months and lasting 3-5 days each. Dust aerosols can harbor a wide diversity of bacteria \[@kelloggCharacterizationAerosolizedBacteria2004\], fungi \[@ramirez-camejoFungalDiversitySahara2022\], virus-like particles \[@griffinAfricanDesertDust\], and minerals \[@formenti2003\]. These aerosols also serve as a significant source of macro and micronutrients such as phosphate, nitrate, and iron \[@millsIronPhosphorusColimit2004; @savoieNonseasaltSulfateNitrate1989 ; @graham1982\] . This addition of otherwise limiting resources can lead to rapid and potentially harmful blooms of certain microbial groups \[@lenesSaharanDustPhosphatic2008; @westrichSaharanDustNutrients2016a ; @westrichSaharanDustNutrients2016a\].

@@

While it is documented that these dust events can stimulate growth of harmful phytoplankton, emerging evidence shows that heterotrophic bacteria are responding to the episodic influx of limiting resources and substrate, resulting in blooms of bacteria that are associated with disease, which we refer to as harmful bacterial blooms. Among those that respond are bacterium belonging to the genus *Vibrio*, a group of ubiquitous marine opportunistic heterotrophs. While considered conditionally rare (comprising of \<1% of the microbial community), under the proper conditions, this genus can rapidly bloom to make up a large percentage of the community over a short period of time [@westrichSaharanDustNutrients2016a ; @westrichVibrioPopulationDynamics2018a]. *Vibrio* species are found in brackish to marine waters with salinities ranging from 0-35 and will grow in waters with temperatures ranging from 5°C-40°C [@sampaioVibrioSppLife2022]. Each year, plumes of Saharan dust travel across the Atlantic via easterly trade winds. These plumes are then deposited in the surface waters of the mid-Atlantic, Caribbean, and Gulf of Mexico. These dust aerosols harbor a wide range of fungi, bacteria, virus-like-particles, minerals, and nutrients (NO3, Fe, PO4). Addition of these nutrients to otherwise oligotrophic settings can result in large and rapid blooms of potentially harmful microbes, presenting a danger to both marine and human health.

@@

Within the *Vibrionaceae* family, there are at least 12 species that are commonly known as human pathogens. Illness can be induced through either a foodborne route via the consumption of raw or uncooked seafood and contaminated drinking water, or through nonfoodborne routes such as wound exposure while swimming. The most common illnesses caused by *Vibrio* include self-limiting diarrhea and cholera (*V. cholerae*), shellfish-induced gastroenteritis (*V. parahaemolyticus, V. vulnificus*), extreme cases of necrotizing fasciitis and septicemia (*V. vulnificus*), and wound and ear infections  (*V. alginolyticus*, *V. vulnificus*) \[@baker-austinVibrioSppInfections2018\]. Some species also pose a threat to marine health, inducing mortality in oyster and clam larvae (*V. tubiashii, V. coralliilyticus*), bacterial bleaching and rapid tissue loss in corals (*V. mediterranei, V. coralliilyticus*) \[@richardsMortalitiesEasternPacific2015; @rosenbergVibrioShiloiOculina2004; @ben-haimVibrioCoralliilyticusSp2003\], and mortality in important aquaculture species such as penaeid shrimp (*V. parahaemolyticus*) and sea breams (*V. harveyi)* \[@tranDeterminationInfectiousNature2013; @haldarIdentificationVibrioHarveyi2010\] among others.

@@

<!--# Have paragraph describing and then have figures. -->

@@

#### Dust AOD and Vibrio Enumeration

@@

Data that is normally distributed was run with Pearson's Correlation Coefficient. Data that is not normally distributed was run with Spearman's Correlation Coefficient.

![Figure 3](images/000081.png)

@@

@@

This study aimed to examine microbial response to Saharan dust deposition in high nutrient coastal waters. Specifically focusing on bacteria belong to the genus *Vibrio*, we found that growth response following dust input varied by site.

@@

### Methods

@@

### Findings/Explanations

<!--# What are AOD units? -->

#### Vibrio and Dust

*Vibrio* growth was found to significantly correlate with dust aerosol optical density at Blind Oso (r = 0.7, p = 0.01), Canals (r = 0.75, p = 0.005), and the Gulf (r = 0.67, p = 0.02). Similar to previous findings peaks in growth were observed 24-48 hours following dust deposition, with both the Gulf and Blind Oso having a peak response 24hr following peak dust deposition, and the Canals 48hr after peak deposition [@westrichVibrioPopulationDynamics2018a; @westrichSaharanDustNutrients2016a]. <!--# Add xfold increases -->.

The introduction of dust constituents have the potential to serve as a temporary alleviation from the limitation of nutrients that are essential for growth. *Vibrio* species have one of the fastest doubling times among bacteria \[@gibsonDistributionBacterialDoubling2018 ; @eagonPSEUDOMONASNATRIEGENSMARINE1962\], and are able to rapidly exploit introduced nutrients due their rapid motility, diverse genomic repertoire, and iron-chelating complexes \[@ringgaardChemotaxisArraysVibrio2018 ; @okadaVibriosCommonlyPossess2005 ; @payneVibrioIronTransport2016\]. These advantageous traits have resulted in *Vibrio* being classified as 'oppurtunitrophs'. The following declines in *Vibrio* gene copies/mL may be attributed to topdown control, such as viral lysis <!--# Cite -->, or bottom-up controls due to an exhaustion of the resources provided by the dust.

#### Site-Specific Dynamics

@@

@@

Similar to previous work done in coastal and offshore sites, the Gulf experienced

Blind Oso Bay is an enclosed shallow tributary that frequently expereince high levels of nutrients (total dissolved nitrogen) and chlorophyll due to its proximity to the Oso Wastewater Plant (OSP). The Bay experienced a dramatic drop in salinity from July 11th to July 19th. This decrease in salinity could be attributed to discharge from the OSP.

The Canals may have a heavily dampened growth response due to a high and constant salinity of 40-42. In general, optimal growth occurs at salinities of \~25-35, and less tolerant species will drop as salinities climb <!--# CITE --> .

Additionally, the Gulf exhibited a sharper increase during the second, heavier dust event, while Blind Oso exhibited a smaller increase during the heavy dust event. This difference in response may be attributed to the nutrient dynamics of the sites. In marine waters, nutrients are typically limiting due to tight microbial recycling <!--# CITE -->. Therefore the introduction on nutrients at such a site may result in rapid uptake and utilization. At Blind Oso, these dust constituents may not be as essential for growth, and the dust may have reached a maximum in how much it can support growth.

@@

[**One important thing to note**]{.ul}: Our data set contains dust data for all of 2022. Our copies_mL data only contains values from 7/7/22 --> 7/19/22. Therefore, these CCA were made with only this time series. This limits our analysis, as we have dust data prior that can explain copies_mL on the earlier dates 7/7 or 7/8. But we cannot incorporate this into the CCA as we have NA for copies_mL for those early days. Below we will run linear models that DO contain these prior dust days. And since we have identified a lag, we can shift the dust data to line up with the copies_mL that it corresponds with directly - and we no longer have NAs. But this ultimately means that the lag identified above. may not be the best fit for our model below, so some additional work is needed that tests other lags in the code.

@@

## Conclusions/Broader Impacts

>>>>

The Sahara Desert is a significant source of atmospheric dust, eliciting an estimated one billion metric tons each year [@dalmeidaModelSaharanDust1986\]. Annually, plumes of this dust travel across the Atlantic via easterly trade winds to be deposited in the surface waters of the Atlantic, Caribbean, and Gulf of Mexico. These dust events are highly episodic, occurring 3- 4 times a year typically in the summer months and lasting 3-5 days each. Dust aerosols can harbor a wide diversity of bacteria \[@kelloggCharacterizationAerosolizedBacteria2004\], fungi \[@ramirez-camejoFungalDiversitySahara2022\], virus-like particles \[@griffinAfricanDesertDust\], and minerals \[@formenti2003\]. These aerosols also serve as a significant source of macro and micronutrients such as phosphate, nitrate, and iron \[@millsIronPhosphorusColimit2004; @savoieNonseasaltSulfateNitrate1989 ; @graham1982\] . This addition of otherwise limiting resources can lead to rapid and potentially harmful blooms of certain microbial groups \[@lenesSaharanDustPhosphatic2008; @westrichSaharanDustNutrients2016a ; @westrichSaharanDustNutrients2016a].

@@

While it is documented that these dust events can stimulate growth of harmful phytoplankton, emerging evidence shows that heterotrophic bacteria are responding to the episodic influx of limiting resources and substrate, resulting in blooms of bacteria that are associated with disease, which we refer to as harmful bacterial blooms. Among those that respond are bacterium belonging to the genus *Vibrio*, a group of ubiquitous marine opportunistic heterotrophs. While considered conditionally rare (comprising of \<1% of the microbial community), under the proper conditions, this genus can rapidly bloom to make up a large percentage of the community over a short period of time [@westrichSaharanDustNutrients2016a ; @westrichVibrioPopulationDynamics2018a]. *Vibrio* species are found in brackish to marine waters with salinities ranging from 0-35 and will grow in waters with temperatures ranging from 5°C-40°C [@sampaioVibrioSppLife2022].

@@

Within the *Vibrionaceae* family, there are at least 12 species that are commonly known as human pathogens. Illness can be induced through either a foodborne route via the consumption of raw or uncooked seafood and contaminated drinking water, or through nonfoodborne routes such as wound exposure while swimming. The most common illnesses caused by *Vibrio* include self-limiting diarrhea and cholera (*V. cholerae*), shellfish-induced gastroenteritis (*V. parahaemolyticus, V. vulnificus*), extreme cases of necrotizing fasciitis and septicemia (*V. vulnificus*), and wound and ear infections  (*V. alginolyticus*, *V. vulnificus*) [@baker-austinVibrioSppInfections2018\]. Some species also pose a threat to marine health, inducing mortality in oyster and clam larvae (*V. tubiashii, V. coralliilyticus*), bacterial bleaching and rapid tissue loss in corals (*V. mediterranei, V. coralliilyticus*) \[@richardsMortalitiesEasternPacific2015; @rosenbergVibrioShiloiOculina2004; @ben-haimVibrioCoralliilyticusSp2003\], and mortality in important aquaculture species such as penaeid shrimp (*V. parahaemolyticus*) and sea breams (*V. harveyi)* \[@tranDeterminationInfectiousNature2013; @haldarIdentificationVibrioHarveyi2010] among others.

@@

@@

#### Dust AOD and Vibrio Enumeration <!--# What are AOD units? -->

@@

Correlation matrices were run on each variable (in order of left to right: temperature, salinity, nitrate+nitrite, orthophosphate, dissolved organic nitrogen, dissolved organic carbon, total dissolved nitrogen, chlorophyll, *Vibrio* gene copies/mL, and dust). A majority of the environmental variables had a non-normal distribution, thus Spearman's Correlation (indicated by the color gradient) was used to examine the relationship between the variables.

![](images/000175.png)

@@

<!--# Add RF graphs here -->

@@

@@

This study aimed to examine microbial response to Saharan dust deposition in high nutrient coastal waters, specifically focusing on bacteria belong to the genus *Vibrio*, a bacterium that is important for marine processes such as carbon cycling, as well as human and marine health.

## Methods

@@

## Findings/Explanations

*Vibrio* growth was found to significantly correlate with dust aerosol optical density at Blind Oso (r = 0.7, p = 0.01), Canals (r = 0.75, p = 0.005), and the Gulf (r = 0.67, p = 0.02). The introduction of dust constituents have the potential to serve as a temporary alleviation from the limitation of nutrients that are essential for growth. *Vibrio* species have one of the fastest doubling times among bacteria [@gibsonDistributionBacterialDoubling2018 ; @eagonPSEUDOMONASNATRIEGENSMARINE1962\], and are able to rapidly exploit introduced nutrients due their rapid motility, diverse genomic repertoire, and iron-chelating complexes \[@ringgaardChemotaxisArraysVibrio2018; @okadaVibriosCommonlyPossess2005 ; @payneVibrioIronTransport2016]. These advantageous traits have resulted in *Vibrio* being classified as 'oppurtunitrophs'. The following declines in *Vibrio* gene copies/mL may be attributed to topdown control, such as viral lysis <!--# Cite -->, or bottom-up controls due to an exhaustion of the resources provided by the dust.

Similar to previous findings, peaks in growth were observed 24-48 hours following dust deposition, with both the Gulf and Blind Oso having a peak response 24hr following peak dust deposition, and the Canals 48hr after peak deposition [@westrichVibrioPopulationDynamics2018a; @westrichSaharanDustNutrients2016a]. <!--# Add xfold increases -->.

@@

*Vibrio* gene copies/mL at the Gulf experienced the greatest change in growth <!--# Add xfold --> as compared to the other two sites. This response in a lower nutrient marine environment lines up with previous work that was conducted in the Florida Keys and North Atlantic, where a 30-fold and 1.6-fold population increase was observed within a 24hr period following dust deposition.

While *Vibrio* growth exhibited a strong correlation at the Canals, the degree of growth was lesser in comparison to the other sites (Figure 1). Field studies have suggested that salinities for optimal growth lie between 5-25, depending on the species [@sullivanEffectsTemperatureSalinity2018]. The higher salinities at this site (40-42) may have attributed to this observed weaker growth response.

*Vibrio* growth was observed at Blind Oso Bay, but slightly dampened in comparison to the Gulf. This site is an enclosed shallow tributary that frequently experiences high levels of nutrients (e.g.total dissolved nitrogen) and chlorophyll due to its proximity to the Oso Wastewater Plant (OSP) [@wetz]. The Bay experienced a dramatic drop in salinity from July 11th to July 19th that could be attributed to discharge from the OSP. While growth was significantly correlated with dust deposition, there are many environmental factors at play at this site, making it difficult to identify the specific mechanisms that are eliciting the growth response.

Additionally, the Gulf exhibited a sharper increase during the second, heavier dust event, while Blind Oso exhibited a smaller increase during the heavy dust event. This difference in response may be attributed to the nutrient dynamics of the sites. In marine waters, nutrients are typically limiting due to tight microbial recycling <!--# CITE -->. Therefore the introduction on nutrients at such a site may result in rapid uptake and utilization. At Blind Oso, a priming effect may be occurring, where these dust constituents may not be as essential for growth following an initial deposition, reaching a maximum in how much it can support growth.

@@

@@

<!--# Difficult to parse out sites  -->

@@

## Conclusions

Ultimately, *Vibrio* growth was observed following dust deposition in each of these sites. However, the degree of response varied by site. This may be explained by each site having a specific and complex environment with varying water chemistry and nutrient levels.

Sites like the Gulf, that have lower nutrient levels, may have a more responsive microbial population, as they are temporarily relieved form nutrient limitation. Inland sites similar to the Canals or Blind Oso may exhibit dampened *Vibrio* growth following dust input, as the dust derived nutrients may be less critical for growth, as there is already a supporting nutrient foundation, or the environmental conditions may not be favorable for growth initially. The growth response at Blind Oso, despite having higher nutrients, suggests that there may be other constiuents in the dust that are not present in the site that elicit this response. More work will be needed to fully understand the dynamics, as there were many other attributes that may have also contributed to growth in the Bay.

To date, our understanding of *Vibrio* and other microbial response to dust input has focused on oligotrophic settings, where dust input is considered to be a critical source of limiting nutrients, and iron in particular. This study aided in filling the gap in how comparable previous findings in oligotrophic settings are to coastal areas with higher ambient nutrient levels. The composition of desert dust can be complex and may have the potential to deliver critical resources that could be more readily exploited by opportunistic bacteria (like *Vibrio*) and elicit a population bloom despite higher baseline nutrient levels. During these bloom periods, human exposure through recreation or consumption of raw seafood may be more likely, presenting a risk to the health of the public.
